---
title: "Write and test a DataRaft adapter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Write and test a DataRaft adapter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

An adapter is a specification with S3 methods. Constructors describe work;
preflight checks configuration; execution performs I/O. Import the generics from
`dataraft.core` and register each method with roxygen `@export` in your package.
Never put credentials, live connections or executable callbacks in `dr_inspect()`.

```{r}
library(dataraft.core)
library(dataraft.adapters)

csv_source <- function(path) structure(list(path = path), class = "example_csv")
dr_check_component.example_csv <- function(x, ...) {
  stopifnot(is.character(x$path), length(x$path) == 1L, file.exists(x$path))
  invisible(x)
}
dr_inspect.example_csv <- function(x, ...) list(type = "CSV", path = x$path)
dr_capabilities.example_csv <- function(x, ...) {
  dr_component_capabilities(read = TRUE, write = FALSE, lazy = FALSE,
    transactions = FALSE, immutable = FALSE)
}
dr_read_source.example_csv <- function(source, ...) {
  dr_check_component(source)
  read.csv(source$path)
}
path <- tempfile(fileext = ".csv")
write.csv(data.frame(id = 1:3), path, row.names = FALSE)
source <- csv_source(path)
dr_test_adapter(source, expected = data.frame(id = 1:3))
unlink(path)
```

For a target implement `dr_write_target(target, data, context, ...)` and return
a descriptive list identifying the committed output. Honour `context$contract`:
validate the stored candidate before exposing it, and fail closed if validation
cannot establish success. A target must never claim success on a partial write.

```{r}
path <- tempfile("adapter-test-")
input <- data.frame(id = 1:3)
result <- dr_test_adapter(dr_target_rds(path), data = input,
  context = list(contract = dr_contract(columns = c(id = "integer"))),
  read_back = function(output) dr_read_source(dr_source_rds(path, output$version)))
unlink(path, recursive = TRUE)
```

Use disposable locations. The helper performs real reads and explicitly requested
writes. Add failure-injection tests for partial writes, rollback, concurrent
writers, invalid contracts, and secrets in exceptions. The helper is a protocol
smoke test, not certification of these backend-specific guarantees.

For OpenLineage, a descriptor can include
`dataset = list(namespace = "postgres://warehouse:5432",
name = "insurance.public.contracts")`. Only provide verified public physical
identities. See the [OpenLineage naming conventions](https://openlineage.io/docs/spec/naming/).

## Adapter conformance

These requirements apply to provider implementations, including built-in ones.

* Constructors and inspection MUST NOT perform I/O or expose credentials.
* Sources MUST return the declared table shape and MUST document snapshot and
  connection ownership. The generic executor freezes lazy output before its gate.
* Writers MUST consume the supplied, checked candidate. They MUST NOT rerun an
  upstream query or silently substitute a different dataset.
* A failed write MUST raise an error. A target claiming transactions MUST leave
  the previously committed state intact, including on a failure during commit.
* Each target MUST document replay behavior: overwrite, append, deduplicate or
  reject. Idempotence MUST NOT be assumed from a successful smoke test.
* Metadata delivery MUST be retryable independently from data publication and
  MUST NOT report a failed data publication as successful.
* Capability declarations MUST describe supported behavior. The core consumes
  `write`, `transactions`, `immutable` and layer configuration in the relevant
  execution paths. Undeclared capabilities are never a guarantee.

`dr_test_adapter()` checks a real roundtrip on disposable resources. Supply
`failure_probe` plus `read_committed` to inject a failure after staging and verify
that committed state remains unchanged. A passing probe establishes that tested
failure only; it does not certify every backend failure or concurrent execution.

### Backend protocols beyond a single source or target

Core delegates provider-owned release operations through S3 methods. Register
methods against the generics exported by `dataraft.core`; do not add a reverse
namespace call to core. These protocols are an extension interface, separate from
the seventeen entry topics in the umbrella reference.

| Protocol | Method arguments after dispatch object | Result and ownership |
| --- | --- | --- |
| `dr_connect_backend` | Provider connection options | Connected backend; caller closes what it opens |
| `dr_close_backend` | Provider close options | Release only provider-owned resources |
| `dr_resolve_release` | Asset ID and optional exact release ID | One release-reference row; missing or ambiguous identities must fail |
| `dr_release_table` | Asset ID and exact release ID | Pinned table reference |
| `dr_read_release_data` | Asset ID and exact release ID | Retained data for that release |
| `dr_registry_data` | Registry table name | Registry data frame |
| `dr_filter_metadata` | Table, asset and run ID filters | Filtered registry rows |
| `dr_configure_target` | Layer settings | Updated target definition, no writes |
| `dr_publish_model_result` | Product, checked model result and expected previous release | Atomically published model result or error |

The other protocols cover source landing, source replacement, model reconstruction,
lineage and metric diagnostics. Their provider methods must retain the same
identity and ownership rules. A method's existence alone does not demonstrate
atomicity; use the conformance and failure probes above on its real backend.
