---
title: "Write and test a DataRaft adapter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Write and test a DataRaft adapter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

An adapter is a specification with S3 methods. Constructors describe work;
preflight checks configuration; execution performs I/O. Import the generics from
`dataraft.core` and register each method with roxygen `@export` in your package.
Never put credentials, live connections or executable callbacks in `dr_inspect()`.

```{r}
library(dataraft.core)
library(dataraft.adapters)

csv_source <- function(path) structure(list(path = path), class = "example_csv")
dr_check_component.example_csv <- function(x, ...) {
  stopifnot(is.character(x$path), length(x$path) == 1L, file.exists(x$path))
  invisible(x)
}
dr_inspect.example_csv <- function(x, ...) list(type = "CSV", path = x$path)
dr_capabilities.example_csv <- function(x, ...) {
  dr_component_capabilities(read = TRUE, write = FALSE, lazy = FALSE,
    transactions = FALSE, partition = FALSE, immutable = FALSE)
}
dr_read_source.example_csv <- function(source, ...) {
  dr_check_component(source)
  read.csv(source$path)
}
path <- tempfile(fileext = ".csv")
write.csv(data.frame(id = 1:3), path, row.names = FALSE)
source <- csv_source(path)
dr_test_adapter(source, expected = data.frame(id = 1:3))
unlink(path)
```

For a target implement `dr_write_target(target, data, context, ...)` and return
a descriptive list identifying the committed output. Honour `context$contract`:
validate the stored candidate before exposing it, and fail closed if validation
cannot establish success. A target must never claim success on a partial write.

```{r}
path <- tempfile("adapter-test-")
input <- data.frame(id = 1:3)
result <- dr_test_adapter(dr_target_rds(path), data = input,
  context = list(contract = dr_contract(columns = c(id = "integer"))),
  read_back = function(output) dr_read_source(dr_source_rds(path, output$version)))
unlink(path, recursive = TRUE)
```

Use disposable locations. The helper performs real reads and explicitly requested
writes. Add failure-injection tests for partial writes, rollback, concurrent
writers, invalid contracts, and secrets in exceptions. The helper is a protocol
smoke test, not certification of these backend-specific guarantees.

## Adapter conformance

The terms MUST and SHOULD below describe the adapter author's contract. They do
not imply that every backend supports transactions or immutable storage.

- Constructors MUST describe work without opening connections or writing data.
  Inspection and error messages MUST omit credentials, access tokens and secret
  connection strings, including when a downstream library raises an error.
- A writer MUST report success only after the output it describes is committed
  according to its documented backend semantics. It MUST NOT return a successful
  descriptor for a partial write or treat missing quality evidence as success.
- An adapter MUST document its atomicity boundary, concurrent-writer behavior,
  retry and idempotency behavior, and what remains after a partial failure.
  Explicitly state when retries append another version or can duplicate data.
  A backend without rollback MUST say so and describe recovery; do not claim
  atomicity merely because a successful call returns one descriptor.
- Capabilities MUST describe supported behavior accurately. They do not supply
  transactions, isolation or idempotency by themselves. Document guarantees that
  depend on filesystem, database or deployment configuration separately.
- Adapters SHOULD provide reproducible failure-injection tests against disposable
  resources for each guarantee they claim. Include failures before commitment,
  after commitment but before response, concurrent attempts and retries where
  relevant. If a phase cannot be tested in ordinary CI, record that boundary.

### A failure-injection test recipe

This small target demonstrates a failure before commitment. It writes staging
data, fails deliberately, and removes its own staging file. The previous output
must survive and DataRaft must retain an error result. In your adapter suite,
replace the injected writer with your actual backend's failure hook and inspect
the real committed object. Use an explicit contract so a missing contract cannot
prevent the writer from being reached and accidentally make this test pass.

```{r}
root <- tempfile("adapter-failure-")
dir.create(root)
committed <- file.path(root, "committed.rds")
staging <- file.path(root, "staging.rds")
previous <- data.frame(id = 1L)
saveRDS(previous, committed)
attempted <- FALSE
failing_writer <- function(data, context) {
  attempted <<- TRUE
  on.exit(unlink(staging), add = TRUE)
  saveRDS(data, staging)
  stop("Injected failure before commit")
}
product <- dr_product("adapter_failure", data.frame(id = 2L)) |>
  dr_add_contract(dr_contract(columns = c(id = "integer"))) |>
  dr_set_target(failing_writer)
failed <- dr_run(product, stop_on_failure = FALSE)
stopifnot(
  attempted,
  identical(failed$status, "error"),
  identical(readRDS(committed), previous),
  !file.exists(staging)
)
unlink(root, recursive = TRUE)
```

An exit handler does not cover a killed process or lost host. Test crash recovery
separately when your adapter claims it. For a non-atomic backend, assert the
documented partial state and recovery procedure instead of assuming rollback.
`dr_test_adapter()` remains a protocol smoke test; neither it nor this synthetic
writer certifies a third-party backend's guarantees.

For OpenLineage, a descriptor can include
`dataset = list(namespace = "postgres://warehouse:5432",
name = "insurance.public.contracts")`. Only provide verified public physical
identities. See the [OpenLineage naming conventions](https://openlineage.io/docs/spec/naming/).
